---
title: "Course Schedule Survey"
subtitle: "Student Sample"
author: "James Van Slyke"
date: today
format:
  docx:
    toc: true
    highlight-style: github
editor: visual
---

```{r}
#| echo: false
#| include: false
#| message: false
library(tidyverse)
library(gt)
library(readr)
CS_Student <- read_csv("/Users/jamesvanslyke/Library/CloudStorage/GoogleDrive-james.vanslyke@gmail.com/My Drive/Current/FPU/Admin/FPU Admin Projects/Course Schedule Surveys/Data/CS_Student.csv")
```

```{r}
#| echo: false
#| include: false
#| message: false
#Dataset Cleaning
#Remove unwanted rows 
CS_Student <- CS_Student %>% 
  slice(-c(1, 2))

#Select the columns that you want 
CS_Student <- CS_Student |> 
  select(Q16:Q12)

#Remove NAs
# Remove rows where column 'A' has NA
CS_Student <- CS_Student |>  
  filter(!is.na(Q16))

```

## Year in School

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  count(Q16) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q16, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() + 
  scale_x_discrete(limits = c("Freshman", "Sophmore", "Junior", "Senior")) +
  labs(x ="")

  
```

## Number of Years at FPU

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  count(Q17) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q17, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() + 
  scale_x_discrete(limits = c("Less than a year", "1 Year", "2 Years", "3 Years", "4 Years", "More than 4 Years")) +
  labs(x ="")


```

## Are you the first person in your family to go to college?

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  count(Q18) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q18, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() +
  labs(x ="")
  
```

## Do you live on campus or do you commute?

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  count(Q2) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q2, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() +
  labs(x ="")

```

## As a commuter, do you prefer coming to campus 2 days a week or 3 days a week?

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  filter(!is.na(Q14)) |> 
  count(Q14) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q14, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() +
  labs(x ="") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(limits = c(0, 60))

```

## How long does it take you to get to campus (in minutes)?

```{r}
#| echo: false 
#| message: false
#| warning: false
df <- CS_Student |> 
  select(Q3) |> 
  drop_na() |> 
  mutate(Q3 = as.numeric(Q3))

df |> 
  drop_na() |> 
  summarise(
    Mean = mean(Q3),
    Median = median(Q3), 
    Standard_Deviation = sd(Q3), 
    Min_value = min(Q3),
    Max_value = max(Q3)
  ) |> 
  gt()
```

## Do you have a job?

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  filter(!is.na(Q4)) |> 
  count(Q4) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q4, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() +
  labs(x ="")

```

## Is your job on campus or off campus?

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  filter(!is.na(Q13)) |> 
  count(Q13) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q13, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() +
  labs(x ="")

```

## How many hours do you work each week?

```{r}
#| echo: false 
#| message: false
#| warning: false
df <- CS_Student |> 
  select(Q5) |> 
  drop_na() |> 
  mutate(Q5 = as.numeric(Q5))

df |> 
  drop_na() |> 
  summarise(
    Mean = mean(Q5),
    Median = median(Q5), 
    Standard_Deviation = sd(Q5), 
    Min_value = min(Q5),
    Max_value = max(Q5)
  ) |> 
  gt()
```

## Which days of the week do you work?

```{r}
#| echo: false 
#| message: false
CS_Student |> 
  select(Q6) |> 
  filter(!is.na(Q6)) |> 
  separate_longer_delim(Q6, delim = ",") |> 
ggplot() +
  geom_bar(aes(x = Q6), fill = "cornflowerblue") +
  theme_minimal() + 
  scale_x_discrete(limits = c("Sunday", "Monday","Tuesday"     ,"Wednesday", "Thursday", "Friday", "Saturday")) +
  labs(x ="", y = "Number")


```

## Are you a FPU Student Athlete?

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  filter(!is.na(Q7)) |> 
  count(Q7) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q7, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() +
  labs(x ="")


```

## Would you be willing to start your school day earlier if it meant you could finish sooner (7:30am to 2:00pm, for example?

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  filter(!is.na(Q8)) |> 
  count(Q8) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q8, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() +
  labs(x ="")

```

## Given the opportunity, would you be interested in taking hybrid courses? 

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  filter(!is.na(Q10)) |> 
  count(Q10) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q10, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() +
  scale_x_discrete(limits =c("Definitely not", "Probably not", "Might or might not", "Probably yes", "Definitely yes")) +
  labs(x ="")

```

## For online classes, do you prefer that they have a zoom component or that it's completely asynchronous?

```{r}
#| echo: false 
#| message: false
Data_Summary <- CS_Student |> 
  filter(!is.na(Q15)) |> 
  count(Q15) |> 
  mutate(Percentage = n /sum(n)*100) 


ggplot(Data_Summary, aes(x = Q15, 
                         y = Percentage)) +
  geom_bar(stat = "identity", 
           fill = "cornflowerblue") +
  geom_text(aes(label = paste0(round(Percentage, 
    1), "%")), vjust = -0.5, size = 4) +
  theme_minimal() +
  labs(x ="")

```

## Approximately how many times have you taken the following courses?

```{r}
#| echo: false
#| message: false
CS_Student |> 
  select(Q11_1:Q11_3) |> 
  rename(Hybrid = Q11_1,
         Online = Q11_2,
         In_person = Q11_3) |> 
  drop_na() |> 
  pivot_longer(cols = everything(), 
        names_to = "variable", values_to = "value") |> 
  mutate(value = as.numeric(value)) |> 
  group_by(variable) |> 
  summarize(n = n(),
            mean = mean(value),
            sd = sd(value),
            se = sd / sqrt(n),
            ci = qt(0.975, df = n - 1) * sd / sqrt(n)) |> 
  ggplot( aes(x = variable,
             y = mean)) +
  theme_minimal() +
  geom_bar(stat = "identity", fill="cornflowerblue") +
  geom_errorbar(aes(ymin=mean-ci,
                    ymax=mean+ci), width=.3, linewidth=1) +
  labs(y = "Mean Frequency", x = "Type of Class") +
  scale_x_discrete(limits = c("In_person", "Online", "Hybrid"))
```

## Rank your preferences for the following courses

```{r}
#| echo: false
#| message: false
CS_Student |> 
  select(Q19_1:Q19_3) |> 
  rename(Hybrid = Q19_1,
         Online = Q19_2,
         In_person = Q19_3) |> 
  drop_na() |> 
  pivot_longer(cols = everything(), 
        names_to = "variable", values_to = "value") |> 
  mutate(value = as.numeric(value)) |> 
  group_by(variable) |> 
  summarize(n = n(),
            mean = mean(value),
            sd = sd(value),
            se = sd / sqrt(n),
            ci = qt(0.975, df = n - 1) * sd / sqrt(n)) |> 
  ggplot( aes(x = variable,
             y = mean)) +
  theme_minimal() +
  geom_bar(stat = "identity", fill="cornflowerblue") +
  geom_errorbar(aes(ymin=mean-ci,
                    ymax=mean+ci), width=.3, linewidth=1) +
  labs(y = "Mean Preference", x = "Type of Class") +
  scale_x_discrete(limits = c("In_person", "Online", "Hybrid"))
```

## Based on the previous question, why do you prefer certain types of courses? 

```{r}
#| echo: false 
#| message: false
library(tidyverse)
library(gt)
library(readr)
CS_Student <- read_csv("/Users/jamesvanslyke/Library/CloudStorage/GoogleDrive-james.vanslyke@gmail.com/My Drive/Current/FPU/Admin/FPU Admin Projects/Course Schedule Surveys/Data/CS_Student.csv")

CS_Student |> 
  select(Q12) |> 
  filter(!is.na(Q12)) |> 
  rename("Comments" = Q12) |> 
  gt() |> 
  tab_options(
    table.width = pct(100), 
    table.align = "center"
  )
```
